---
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#Modules
library(readxl)
library(openxlsx)
library(astsa)
library(dynlm) #Time series regression 
library(broom) #LM test, table presentations
library(FinTS) #For ARCH test
library(forecast)
library(tseries)
library(TTR)
library(TSA)
library(graphics)
library(portes)
library(tseries) #For unit root test
library(car) #For robust standard errors
library(rugarch) #For Garch Models
library(vars) #For using VAR
library(nlWaldTest) # For testing non-linear wald test
library(lmtest) #For BP test
library(ggplot2)
library(tsibble)
library(tseries)
library(MASS)
library(forecast)
library(aTSA)
library(grid)
library(gridExtra)
library(magrittr)
library(knitr)
library(sandwich)
library(tsbox)
library(stats)
library(zoo)
library(vrtest)
library(fGarch)
library(aTSA)
library(FinTS) 
library(lmtest)
library(forecast)
library(TSA)
library(tseries)
library(xts)
library(readxl)
library(tidyverse)
```

```{r}
#Path and dataset
path = "D:/Semester 5/MPDW/Project/"
df_average <- read.csv(paste(path, "Rata Harian.csv", sep=""))[,2:3]
```

```{r}
#pakai 15 bulan
Bulan=15
temp=-(Bulan*30-1)+length(df_average$Rataan.Ketinggian.Air.Laut)
df=df_average[temp:length(df_average$Rataan.Ketinggian.Air.Laut),]
df["Tanggal"]=seq(1,length(df$Tanggal))
colnames(df)= c("Periode", "Rataan.Ketinggian.Air.Laut")
df.ts=ts(df$Rataan.Ketinggian.Air.Laut)
```

```{r}
#Train Test Split
train=df[1:360,]
train.ts=ts(train$Rataan.Ketinggian.Air.Laut)
temp=length(df$Periode)
test=df[361:temp,]
test.ts=ts(test$Rataan.Ketinggian.Air.Laut)
```

```{r}
#Plotting Train and Test
ggplot() +
  geom_line(data = test, aes(x = Periode, y = Rataan.Ketinggian.Air.Laut, col = "Test")) +
   geom_line(data = train, aes(x = Periode, y = Rataan.Ketinggian.Air.Laut, col = "Train"))+ labs(x = "Periode Waktu", y = "Tinggi Air", color = "Legend") +
  scale_colour_manual(name="legend:", breaks = c("Train", "Test"),
                      values = c("blue", "red")) +
  theme_bw() + theme(legend.position = "bottom", plot.caption = element_text(hjust=0.5, size=12))
```

#Uji stasioneritas

```{r}
#Dengan plot
acf(train.ts , lag.max=30, main = "ACF Plot") #tidak stasioner dlm rataan
```

```{r}
#Uji Formal
adf.test(train.ts)
```

#Box cox

```{r}
n = length(train.ts)
index <- seq(1:n) #Sampai periode terakhir data train
bc = boxcox(train.ts~index, lambda = seq(0,6,by=1))
#Nilai Rounded Lambda
lambda <- bc$x[which.max(bc$y)]
best.lambda = bc$x[which(bc$y==max(bc$y))]
best.lambda
#SK
bc$x[bc$y > max(bc$y) - 1/2 * qchisq(.95,1)]
```

#Penanganan ##BOX COX, DIFF

```{r}
 # d <- 1
 # diff_train_data <- diff(train.ts, differences = d)
# lambda <- BoxCox.lambda(train.ts)
# transformed_train_data <- BoxCox(train.ts, lambda)

# dif_ori <- diff(transformed_train_data, differences = 1)

dif_ori <- diff(train.ts, differences = 1)
```

```{r}
acf(dif_ori, main = "ACF Plot Differencing 1x")
```

```{r}
plot.ts(dif_ori, xlab='Time Period', ylab='1st Difference', main="Differencing 1x Tinggi Permukaan Air data Train")
```

#Model Terbaik

```{r}
acf(dif_ori, lag.max=30) #ARIMA(0,1,1)
pacf(dif_ori, lag.max=30) #ARIMA(1,1,0) ARIMA(4,1,0)
eacf(dif_ori) #ARIMA(0,1,1), ARIMA(0,1,2), ARIMA(1,1,2)
```

# ARIMA(p,d,q)

```{r}
modelARIMA <- Arima(train.ts, order=c(1,1,3), method = "ML")
summary(modelARIMA) #6035, 6041, 6035.9,6035.32, 6021.4 
lmtest::coeftest(modelARIMA)
```

# ARIMA(1,1,2)

```{r}
model2.of <- Arima(train.ts, order=c(1,1,2),method="ML")
summary(model2.of) #AIC=2277.54 scale to 6021.4

#Signifikansi koefisien
lmtest::coeftest(model2.of) #Seluruh parameter signifikan
tsdiag(model2.of) #sisaan terlihat acak dan bebas
checkresiduals(model2.of)
autoplot(model2.of) #model stable, tidak ada titik yang diluar
```

##Get predicted train and test

```{r}
# train_pred <- predict(model2.of, n.ahead = length(train$Rataan.Ketinggian.Air.Laut))$pred
# 
# test_pred <- predict(model2.of, n.ahead = length(test$Rataan.Ketinggian.Air.Laut))$pred


ramalan.da <- forecast::forecast(model2.of, h = 90) 
ramalan.da
data.ramalan.da <- ramalan.da$mean
plot(ramalan.da)

pt_1 <- train.ts[360] #nilai akhir data latih
hasil.forc.Diff <- data.ramalan.da
hasil <- diffinv(hasil.forc.Diff, differences = 1) + pt_1
plot(hasil)

#has.1 sama hasilnta dengan: cumsum(c(pt_1,hasil.forc.Diff))
ts.plot(train.ts,hasil.balik.trans)


test_pred <- predict(model2.of, n.ahead = length(test$Rataan.Ketinggian.Air.Laut))$pred

perbandingan <- rbind(test.ts,test_pred)
```

# Highliht3

```{r}
# Simulated data
actual_values <- test.ts
predicted_values <- as.matrix(test_pred)

# Calculate RMSE
rmse <- sqrt(mean((actual_values - predicted_values)^2))

# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100


# Create an evaluation matrix
evaluation_matrix <- data.frame(
  RMSE = rmse,
  MAPE = mape
)

# Print the evaluation matrix
print(evaluation_matrix)
```

# Diagnostik Model

```{r}
#Eksplorasi dengan plot
sisaan.da <- model2.of$residuals 
par(mfrow=c(2,2)) 
qqnorm(sisaan.da) 
qqline(sisaan.da, col = "blue", lwd = 2) 
plot(c(1:length(sisaan.da)),sisaan.da) 
acf(sisaan.da) 
pacf(sisaan.da) 
par(mfrow = c(1,1))
```

```{r}
#Uji Formal
#1) Sisaan Menyebar Normal
shapiro.test(sisaan.da)
ks.test(sisaan.da,"pnorm")
jarque.bera.test(sisaan.da) #Sisaan tidak menyebar normal
  
#2) Sisaan saling bebas/tidak ada autokorelasi
Box.test(sisaan.da, type = "Ljung")  #tak tolak H0 > sisaan saling bebas

#3) Sisaan homogen
Box.test((sisaan.da)^2, type = "Ljung")  #Tolak H0 > Sisaan heterogen

#4) Nilai tengah sisaan sama dengan nol
t.test(sisaan.da, mu = 0, conf.level = 0.95)  #tak tolak h0 > nilai tengah sisaan sama dengan 0
```

$H_0$ : Sisaan menyebar normal $H_1$ : Sisaan tidak menyebar normal
Tolak, Sisaan tidak menyebar normal (terlanggar)

$H_0$ : Sisaan saling bebas $H_1$ : Sisaan tidak tidak saling bebas Tak
tolak, Sisaan saling bebas (terpenuhi)

$H_0$ : Ragam sisaan homogen $H_1$ : Ragam sisaan tidak homogen Tolak,
Sisaan tidak homogen (terlanggar)

$H_0$ : nilai tengah sisaan sama dengan 0 $H_1$ : nilai tengah sisaan
tidak sama dengan 0 Tak tolak, Nilai tengah sisaan sama dengan nol
(terpenuhi)

# ACRH-GARCH

## Uji sisaan Mean model

```{r}
sisaan <- model2.of$residuals
e_topi <- ts(sisaan)
plot.ts(e_topi, main = "Plot Sisaan")

e_topisq <- ts(sisaan^2)
plot.ts(e_topisq)
```

```{r}
acf(e_topi, main = "Plot ACF")
pacf(e_topi, main = "Plot PACF")
```

```{r}
acf(e_topisq) 
```

## Uji Efek ARCH

```{r}
for (i in 1:10) {
  ArchTest <- ArchTest(sisaan.da, lags=i, demean=TRUE)
  cat("P Value LM Test lag ke", i,"adalah" , ArchTest$p.value, "\n") }
```

P-val signifikan hanya di lag 1-7, menunjukkan bahwa asumsi
heteroskedastisitas ARCH signifikan.

```{r}
#Menentukan model volatilitas jika efek ARCH secara statistik signifikan
#Pendugaan model ARCH/GARCH dengan function garch secara simultan
marina.arch1 <- garch(train.ts,c(1,2)) #ordo ARCH(1)
```

```{r}
mariniarch1 <- summary(marina.arch1)
marinaarch1
```

hypotheses for the Jarque-Bera test in the context of a GARCH model: H0:
The residuals (or standardized residuals) from the GARCH model follow a
normal distribution.

H1: The residuals (or standardized residuals) from the GARCH model do
not follow a normal distribution. -\> THIS

Box-Ljung = theres autocol in lags (lag tdk saling bebas), true evidence
autocorrelation in the residuals

```{r}
#str(sbydarch)
plot(marina.arch1) 
```

```{r}
hist(marinaarch1$residuals)
```

```{r}
qqnorm(marinaarch1$residuals)
qqline(marinaarch1$residuals, col = "red", lwd = 2)
```

## Model GARCH

```{r}
# Contoh spesifikasi model ARCH(1)
spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 0)),
                   mean.model = list(armaOrder = c(0, 0)))

# Estimasi model ARCH
arch_model <- ugarchfit(spec, data = your_data_here)

```

### GARCH(1,1)

```{r}
garchSpec11 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit11 <- ugarchfit(spec=garchSpec11, data=diff_train_data)
coef(garchFit11)
```

```{r}
garchFit11 #AIC= 6.2693, sign. not sig. 
```

### GARCH(1,2)

```{r}
garchSpec12 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,2)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit12 <- ugarchfit(spec=garchSpec12, data=diff_train_data)
coef(garchFit12)
```

```{r}
garchFit12 #AIC=6.2606, sign not sig. 
```

### GARCH(1,3)

```{r}
garchSpec13 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,3)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit13 <- ugarchfit(spec=garchSpec13, data=diff_train_data)
coef(garchFit13)
```

```{r}
garchFit13 #AIC=6.2661,bias not sig.
```

### GARCH(1,4)

```{r}
garchSpec14 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,4)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit14 <- ugarchfit(spec=garchSpec14, data=diff_train_data)
coef(garchFit14)
```

```{r}
garchFit14 #AIC=6.2788, bias not sig.
```

GARCH(1,2)

### GARCH(2,1)

```{r}
garchSpec21 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(2,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit21 <- ugarchfit(spec=garchSpec21, data=diff_train_data)
coef(garchFit21)
```

```{r}
garchFit21 #AIC=6.2749, bias not sig.
```

### GARCH(2,2)

```{r}
garchSpec22 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(2,2)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit22 <- ugarchfit(spec=garchSpec22, data=diff_train_data)
coef(garchFit22)
```

```{r}
garchFit22 #AIC=6.2661
```

### GARCH(2,3)

```{r}
garchSpec23 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(2,3)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit23 <- ugarchfit(spec=garchSpec23, data=diff_train_data)
coef(garchFit23)
```

```{r}
garchFit23 #AIC=6.2717
```

### GARCH(2,4)

```{r}
garchSpec24 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(2,4)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit24 <- ugarchfit(spec=garchSpec24, data=diff_train_data)
coef(garchFit24)
```

```{r}
garchFit24
```

### GARCH(3,1)

```{r}
garchSpec31 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(3,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit31 <- ugarchfit(spec=garchSpec31, data=diff_train_data)
coef(garchFit31)
```

```{r}
garchFit31 #AIC=6.2576
```

### GARCH(3,2)

```{r}
garchSpec32 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(3,2)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit32 <- ugarchfit(spec=garchSpec32, data=diff_train_data)
coef(garchFit32)
```

```{r}
garchFit32 #AIC=6.2623
```

### GARCH(3,3)

```{r}
garchSpec33 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(3,3)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit33 <- ugarchfit(spec=garchSpec33, data=diff_train_data)
coef(garchFit33)
```

```{r}
garchFit33 #AIC=6.2679
```

### GARCH(4,1)

```{r}
garchSpec41 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(4,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit41 <- ugarchfit(spec=garchSpec41, data=diff_train_data)
coef(garchFit41)
```

```{r}
garchFit41 #AIC=6.2621
```

### GARCH(5,1)

```{r}
garchSpec51 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(5,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit51 <- ugarchfit(spec=garchSpec51, data=diff_train_data)
coef(garchFit51)
```

```{r}
garchFit51 #AIC=6.2652
```

### GARCH(6,1)

```{r}
garchSpec61 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(6,1)),
                          mean.model=list(armaOrder=c(1,2)))
garchFit61 <- ugarchfit(spec=garchSpec61, data=diff_train_data)
coef(garchFit61)
```

```{r}
garchFit61 #AIC=6.3137
```

### ARCH(1)
```{r}
arch1 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,0)),
                          mean.model=list(armaOrder=c(1,2)))
archFit1 <- ugarchfit(spec=arch1, data=diff_train_data)
coef(archFit1)

archFit1
```
# Forecast ARCH(1)
```{r}
forecast_arch <- ugarchforecast(archFit1, diff_train_data, 90)
plot(forecast_arch, which=1)

pt_2 <- train.ts[360]
hasil.forc.Diff2 <- forecast_arch@forecast[["seriesFor"]]
hasil2 <- diffinv(hasil.forc.Diff2, differences = 1) + pt_2
hasil2

perbandingan_arch <- data.frame("Aktual" = test.ts,
                           "Ramalan" = hasil2[-1])


forecast::accuracy(ts(hasil2[-1]), ts(test.ts))
```

```{r}
# Simulated data
actual_values <- test.ts
predicted_values <- hasil2[-1]

# Calculate RMSE
rmse <- sqrt(mean((actual_values - predicted_values)^2))

# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100


# Create an evaluation matrix
evaluation_matrix <- data.frame(
  RMSE = rmse,
  MAPE = mape
)

# Print the evaluation matrix
print(evaluation_matrix)
```

# Final Plot
```{r}
gaktaumales=data.frame(periode= seq(length(train$Periode),450),
           predicted=c(train$Rataan.Ketinggian.Air.Laut[length(train$Rataan.Ketinggian.Air.Laut)], predicted_values))

ggplot() +
  geom_line(data = train, aes(x = Periode, y = Rataan.Ketinggian.Air.Laut, col = "Train"))+
  geom_line(data = gaktaumales, aes(x = periode, y = predicted, col = "Forecasted")) +
  scale_colour_manual(name="legend:", breaks = c("Train", "Forecasted"),
                      values = c("blue", "red")) +
  theme_bw() + theme(legend.position = "bottom", plot.caption = element_text(hjust=0.5, size=12))
```


### ARCH(2)
```{r}
arch2 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(2,0)),
                          mean.model=list(armaOrder=c(1,2)))
archFit2 <- ugarchfit(spec=arch2, data=diff_train_data)
coef(archFit2)

archFit2
```

## Uji Diagnostik GARCH(0,1)

```{r}
marina.garch12 <- garch(train.ts,c(0,1)) #ordo ARCH(1)
```

```{r}
marinagarch12 <- summary(marina.garch12)
marinagarch12
```

## Uji Diganostik GARCH(3,1)

```{r}
marina.garch31 <- garch(train.ts,c(3,1)) #ordo ARCH(3)
```

```{r}
marinagarch31 <- summary(marina.garch31)
marinagarch31
```

```{r}
#str(sbydarch)
plot(marina.garch12) 
```

## Peramalan

```{r}
## ramalan dari data train
model <-Arima(train.ts,order=c(1,1,2))
ramalan_model<-forecast::forecast(model,h=90)
plot(ramalan_model)
```

```{r}
train.ts
test.ts
```

```{r}
hasil_perbandingan<- as.data.frame(cbind(test.ts,ramalan_model$fitted))
hasil_perbandingan
```

# Highlight

```{r}
actual_values <- test.ts
predicted_values <- ramalan_model$fitted

# Evaluation metrics
mae <- mean(abs(actual_values - predicted_values))
mse <- mean((actual_values - predicted_values)^2)
rmse <- sqrt(mse)
# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100

# Calculate residuals
residuals <- actual_values - predicted_values

# Calculate AIC and BIC
n <- length(actual_values)
k <- 2  # Number of parameters in the model (change accordingly)
aic <- n * log(mse) + 2 * k
bic <- n * log(mse) + k * log(n)

# Calculate Mean Squared Deviation (MSD)
msd <- mean(residuals^2)

# Calculate Mean Absolute Deviation (MAD)
mad <- mean(abs(residuals))

# Create an evaluation matrix
evaluation_matrix <- data.frame(
  MAE = mae,
  MSE = mse,
  RMSE = rmse,
  MAPE = mape,
  AIC = aic,
  BIC = bic,
  MSD = msd,
  MAD = mad
)

# Print the evaluation matrix
print(evaluation_matrix)
```

### Model GARCH(1,2)

```{r}
# GARCH(1,2)
m.12 = garch(diff_train_data,order=c(1,2),trace =FALSE)
summary(m.12)
```

### ARIMA(1,1,3)-GARCH(1,2)

```{r}
#ARIMA(1,1,3)+GARCH(1,2)
model12<-ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,2)), 
                  mean.model = list(armaOrder = c(1, 3), include.mean = TRUE), 
                  distribution.model = "norm")
                  
m.12<-ugarchfit(spec=model12,data=diff_train_data, out.sample = 100)
plot(m.12,which="all")
```

```{r}
m.12
```

```{r}
forc = ugarchforecast(m.12, data = diff_train_data, n.ahead = 90, n.roll =10)
print(forc)
```

```{r}
plot(forc, which= "all")
```

```{r}
garchSpec12 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,2)),
                          mean.model=list(armaOrder=c(1,2)))

garch_fit <- ugarchfit(spec = garchSpec12, data = train.ts) 

garch_forecast <- ugarchforecast(garch_fit, n.ahead = 90)  # Change '10' to the number of periods you want to forecast
garch12forecast <- garch_forecast@forecast$seriesFor


hasil_perbandingan<- as.data.frame(cbind(test.ts,garch12forecast))
hasil_perbandingan
```

#Highlight2

```{r}
actual_values <- test.ts
predicted_values <- garch12forecast

# Evaluation metrics
mae <- mean(abs(actual_values - predicted_values))
mse <- mean((actual_values - predicted_values)^2)
rmse <- sqrt(mse)
# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100

# Calculate residuals
residuals <- actual_values - predicted_values

# Calculate AIC and BIC
n <- length(actual_values)
k <- 2  # Number of parameters in the model (change accordingly)
aic <- n * log(mse) + 2 * k
bic <- n * log(mse) + k * log(n)

# Calculate Mean Squared Deviation (MSD)
msd <- mean(residuals^2)

# Calculate Mean Absolute Deviation (MAD)
mad <- mean(abs(residuals))

# Create an evaluation matrix
evaluation_matrix <- data.frame(
  MAE = mae,
  MSE = mse,
  RMSE = rmse,
  MAPE = mape,
  AIC = aic,
  BIC = bic,
  MSD = msd,
  MAD = mad
)

# Print the evaluation matrix
print(evaluation_matrix)
```

```{r}
garchSpec13 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,3)),
                          mean.model=list(armaOrder=c(1,2)))

garch_fit <- ugarchfit(spec = garchSpec13, data = train.ts) 

garch_forecast <- ugarchforecast(garch_fit, n.ahead = 90)  # Change '10' to the number of periods you want to forecast
garch13forecast <- garch_forecast@forecast$seriesFor


hasil_perbandingan<- as.data.frame(cbind(test.ts,garch13forecast))
hasil_perbandingan
```

```{r}
actual_values <- test.ts
predicted_values <- garch13forecast

# Evaluation metrics
mae <- mean(abs(actual_values - predicted_values))
mse <- mean((actual_values - predicted_values)^2)
rmse <- sqrt(mse)
# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100

# Calculate residuals
residuals <- actual_values - predicted_values

# Calculate AIC and BIC
n <- length(actual_values)
k <- 2  # Number of parameters in the model (change accordingly)
aic <- n * log(mse) + 2 * k
bic <- n * log(mse) + k * log(n)

# Calculate Mean Squared Deviation (MSD)
msd <- mean(residuals^2)

# Calculate Mean Absolute Deviation (MAD)
mad <- mean(abs(residuals))

# Create an evaluation matrix
evaluation_matrix <- data.frame(
  MAE = mae,
  MSE = mse,
  RMSE = rmse,
  MAPE = mape,
  AIC = aic,
  BIC = bic,
  MSD = msd,
  MAD = mad
)

# Print the evaluation matrix
print(evaluation_matrix)
```

```{r}
garchSpec13 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(1,3)),
                          mean.model=list(armaOrder=c(1,2)))

garch_fit <- ugarchfit(spec = garchSpec13, data = diff_train_data) 

garch_forecast <- ugarchforecast(garch_fit, n.ahead = 90)  # Change '10' to the number of periods you want to forecast
garch13forecast <- garch_forecast@forecast$seriesFor

# Perform inverse differencing to obtain forecasts in the original scale
garch13_forecast_original <- cumsum(c(tail(train.ts, 1), garch13forecast)) 

# Combine train and forecasted data
all_data <- c(train.ts, garch13_forecast_original)

# Print the resulting data (train + forecasted values)
print(all_data)

# Compare with test data
comparison_df <- data.frame(test.ts = test.ts, garch13_forecast_original = tail(garch13_forecast_original, 90))

# Print the comparison data frame
print(comparison_df)
```

```{r}
actual_values <- test$Rataan.Ketinggian.Air.Laut
predicted_values <- garch13_forecast_original

# Evaluation metrics
mae <- mean(abs(actual_values - predicted_values))
mse <- mean((actual_values - predicted_values)^2)
rmse <- sqrt(mse)
# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100

# Calculate residuals
residuals <- actual_values - predicted_values

# Calculate AIC and BIC
n <- length(actual_values)
k <- 2  # Number of parameters in the model (change accordingly)
aic <- n * log(mse) + 2 * k
bic <- n * log(mse) + k * log(n)

# Calculate Mean Squared Deviation (MSD)
msd <- mean(residuals^2)

# Calculate Mean Absolute Deviation (MAD)
mad <- mean(abs(residuals))

# Create an evaluation matrix
evaluation_matrix <- data.frame(
  MAE = mae,
  MSE = mse,
  RMSE = rmse,
  MAPE = mape,
  AIC = aic,
  BIC = bic,
  MSD = msd,
  MAD = mad
)

# Print the evaluation matrix
print(evaluation_matrix)
```

```{r}
garchSpec31 <- ugarchspec(variance.model=list(model="sGARCH", garchOrder=c(3,1)),
                          mean.model=list(armaOrder=c(1,2)))

garch_fit <- ugarchfit(spec = garchSpec31, data = diff_train_data) 

garch_forecast <- ugarchforecast(garch_fit, n.ahead = 90)  # Change '10' to the number of periods you want to forecast
garch31forecast_diff <- garch_forecast@forecast$seriesFor


# Perform inverse differencing to obtain forecasts in the original scale
garch31_forecast_original <- cumsum(c(tail(train.ts, 1), garch31forecast_diff)) 

# Combine train and forecasted data
all_data <- c(train.ts, garch31_forecast_original)

# Create a data frame for comparison
hasil_perbandingan <- data.frame(test.ts = test.ts, garch31_forecast_original = tail(garch31_forecast_original, 90))

# Print the resulting data (train + forecasted values)
print(all_data)

# Print the comparison data frame
print(hasil_perbandingan)
```

```{r}
actual_values <- test$Rataan.Ketinggian.Air.Laut
predicted_values <- garch31_forecast_original

# Evaluation metrics
mae <- mean(abs(actual_values - predicted_values))
mse <- mean((actual_values - predicted_values)^2)
rmse <- sqrt(mse)
# Calculate MAPE
mape <- mean(abs((actual_values - predicted_values) / actual_values)) * 100

# Calculate residuals
residuals <- actual_values - predicted_values

# Calculate AIC and BIC
n <- length(actual_values)
k <- 2  # Number of parameters in the model (change accordingly)
aic <- n * log(mse) + 2 * k
bic <- n * log(mse) + k * log(n)

# Calculate Mean Squared Deviation (MSD)
msd <- mean(residuals^2)

# Calculate Mean Absolute Deviation (MAD)
mad <- mean(abs(residuals))

# Create an evaluation matrix
evaluation_matrix <- data.frame(
  MAE = mae,
  MSE = mse,
  RMSE = rmse,
  MAPE = mape,
  AIC = aic,
  BIC = bic,
  MSD = msd,
  MAD = mad
)

# Print the evaluation matrix
print(evaluation_matrix)
```



